---
title: "Creating packages really quickly with usethis (and devtools)"
output: html_document
date: "2024-07-28"
---

## Introduction

The usethis package streamlines the setup and maintenance of R packages by automating repetitive tasks. This cheatsheet demonstrates how to use usethis and related packages to create a base package, develop functions, write function tests, add licenses, configure the package for GitHub, and manage dependencies.

## Setup

```{r}
# Install the necessary packages
install.packages("usethis")
install.packages("devtools")
```

## 1: Initialise the package and git

```{r}
# To create a new package, use the `create_package()` function. This will create a new directory with the basic structure of an R package.
# usethis::create_package("randomizer")

# The use_git() function initializes a Git repository in your package directory and adds common files to `.gitignore`.
usethis::use_git()

# You may need to create a GitHub token using create_github_token()
# usethis::create_github_token()

# To connect your package to a GitHub repository, use the `use_github()` function. This will create a new GitHub repository and link it to your local Git repository.
usethis::use_github()

```
## 2: Creating key files

### 2.a: readme

```{r}

# This use_description() function creates a `DESCRIPTION` file for your package which contains metadata about your package, such as its title, description, version, author, and dependencies.
usethis::use_description(fields = list(
  Description = "An example package for generating random numbers.",
  
  `Authors@R` = c(
    person(given = "Dylan", family = "Carbone", email = "dylcar@ceh.ac.uk", role = c("aut", "cre", "cph")),
    person(given = "Alice", family = "Smith", email = "alice.smith@example.com", role = c("aut")),
    person(given = "Bob", family = "Johnson", email = "bob.johnson@example.com", role = c("ctb")),
    person(given = "Carol", family = "Williams", email = "carol.williams@example.com", role = c("ctb"))
  )
))

# The use_mit_license() function adds an MIT license to your package. It creates a `LICENSE` file and populates it with the MIT license text.
usethis::use_mit_license("Dylan Carbone")

```

### 2.b: functions

```{r}

# The `use_r()` function creates a new R script for your function in the `R` directory of your package.
usethis::use_r("random_numbers")

# We'll write an example function with roxygen documentation
writeLines({'
#\' Generate Random Numbers
#\'
#\' This function generates a specified number of random numbers within a given range.
#\' It uses `dplyr::filter` to ensure the numbers are within the specified range.
#\'
#\' @param n A numeric value indicating the number of random numbers to generate.
#\' @param min A numeric value indicating the minimum value of the range.
#\' @param max A numeric value indicating the maximum value of the range.
#\' @return A numeric vector of random numbers.
#\' @import dplyr
#\' @examples
#\' random_numbers(5, 1, 10)
#\' random_numbers(10, 0, 1)
#\' @export
random_numbers <- function(n, min, max) {
  if (!is.numeric(n) || !is.numeric(min) || !is.numeric(max)) {
    stop("All arguments must be numeric values.")
  }
  if (n <= 0) {
    stop("The number of random numbers to generate must be positive.")
  }
  if (min > max) {
    stop("The minimum value must be less than or equal to the maximum value.")
  }
  
  # Generate random numbers
  random_numbers <- runif(n, min, max)
  
  # Create a data frame
  random_df <- data.frame(numbers = random_numbers)
  
  # Use dplyr::filter to keep numbers within the range (though they already are)
  filtered_df <- dplyr::filter(random_df, numbers >= min & numbers <= max)
  
  # Return the filtered numbers as a vector
  return(filtered_df$numbers)
}
'}, "R/random_numbers.R")

```

### 2.c: vignettes

```{r}
# The use_vignette() function writes a vignette to be filled with detailed examples and explanations of how to use your package.
usethis::use_vignette("vignette", "Generating random numbers with randomizer")

# We'll write an example vignette
writeLines({
  '---
title: "Generating random numbers with randomizer"
output: rmarkdown::html_vignette
vignette: >
  %\\VignetteIndexEntry{Generating random numbers with randomizer}
  %\\VignetteEngine{knitr::rmarkdown}
  %\\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(randomizer)
library(ggplot2)

set.seed(123)
random_data <- random_numbers(100, 50, 50)
qplot(random_data, bins=30, main="Histogram of Random Numbers")

```'}, "vignettes/vignette.Rmd")
```

### 2.d: readme

```{r}
# The use_readme_rmd() and use_readme_md() functions generate a 'README.Rmd' and 'README.md file respectively'.
# usethis::use_readme_rmd()
usethis::use_readme_md()
```

## 3: dependencies

```{r}

# You can add dplyr to the suggests field
usethis::use_package("dplyr", type = "Imports")
# You can add ggplot2 to the suggests field
usethis::use_package("ggplot2", type = "Suggests")

# Adds a dependency to the Remotes field in DESCRIPTION.
# This is useful when you want to specify a development version of a package hosted on GitHub.
# usethis::use_dev_package("username/repository")

```

## 4: Package data

```{r}
# Use use_data_raw() to create a script dedicated to generating data to be bundled with the package
usethis::use_data_raw("my_data", open = FALSE)

# Write the script to generate the data frame and save it in data-raw/my_data.R
writeLines({'
# Script to create and save the my_data dataset

# Create the data frame
my_data <- data.frame(
  id = 1:5,
  value = c(10, 20, 30, 40, 50)
)

# The use_data() function bundles the data with the package
usethis::use_data(my_data, overwrite = TRUE)
'}, "data-raw/my_data.R")

# To actually generate and save the data, source the script
source("data-raw/my_data.R")

```

## 5: Testing

```{r}
# The `use_testthat()` function sets up the testing infrastructure with the `testthat` package.
# This function creates the `tests/testthat` directory and adds the necessary files for testing.
usethis::use_testthat()

# Creating tests for the function
# Use the `use_test()` function to create a new test file.
# This function creates a test script in the `tests/testthat` directory of your package.
usethis::use_test("random_numbers")

# Write your test code in the newly created `tests/testthat/test-random_numbers.R` file.
writeLines({'
library(testthat)
library(randomizer)

test_that("random_numbers generates correct number of values", {
  set.seed(123)
  result <- random_numbers(5, 1, 10)
  expect_length(result, 5)
})

test_that("random_numbers generates values within the specified range", {
  set.seed(123)
  result <- random_numbers(5, 1, 10)
  expect_true(all(result >= 1 & result <= 10))
})

test_that("random_numbers handles invalid inputs", {
  expect_error(random_numbers(-5, 1, 10), "positive")
  expect_error(random_numbers(5, 10, 1), "minimum value must be less than or equal to the maximum value")
  expect_error(random_numbers("five", 1, 10), "numeric values")
})
'}, "tests/testthat/test-random_numbers.R")
```

## 6: Continuous integration

```{r}
# The use_github_action function configures GitHub Actions for R CMD check using a standard template.
usethis::use_github_action("check-standard")

# The use_github_action() function creates a GitHub Actions workflow file that runs test coverage checks on your package.
usethis::use_github_action("test-coverage")

```

## 7: Package rendering, documenting, testing, and pushing with devtools

```{r}

# Re-generate documentation with `devtools::document()`. This function uses `roxygen2` to generate the documentation files (e.g., `NAMESPACE`, `.Rd` files) from the inline comments in your R scripts.
devtools::document()

# Install your package locally with `devtools::install()`.
devtools::install()

# Render the vignette with build_vignettes()
devtools::build_vignettes()

# Use devtools render() to convert the `README.Rmd` file to `README.md`.
# rmarkdown::render("README.Rmd")

# Use devtools check() to run a comprehensive check of your package locally.
devtools::check()

```

##9: Other cool functions

```{r}

# use_version() updates to the next patch, minor, or major version (e.g., 1.0.0 to 1.0.1).
usethis::use_version("patch")
usethis::use_version("minor")
usethis::use_version("major")

# use_citation() generates a citation to encourage users of your package to cite an article or book.
usethis::use_citation()

# use_news_md() generates a news md file to record changes made in each version of the package, such as new features, bug fixes, and other updates.
usethis::use_news_md()

# use_code_of_conduct() adds a code of conduct to your package by creating a `CODE_OF_CONDUCT.md` file with a template code of conduct.
usethis::use_code_of_conduct("dylcar@ceh.ac.uk")

# browse_package() opens the CRAN package webpage in the default browser for quickly accessing documentation and information about a specific package.
usethis::browse_package(package = "dplyr")

# use_git_hook() adds Git hooks to your repository to enforce code standards or run tests before pushing changes, such as adding a pre-commit hook to run lintr on staged files.
usethis::use_git_hook("pre-commit", "R -e 'lintr::lint_dir()'")

# use_tidy_github() configures your repository to align with tidyverse conventions, including settings for GitHub Actions, issue templates, and labels to ensure consistency across tidyverse packages.
usethis::use_tidy_github()

# use_pkgdown_github_pages() sets up a GitHub Actions workflow to run `pkgdown` to automate the process of building and deploying a package website using `pkgdown`.
usethis::use_pkgdown_github_pages()

# use_github_links() adds GitHub links to DESCRIPTION by adding URLs for the BugReports and URL fields in the DESCRIPTION file.
usethis::use_github_links(overwrite = TRUE)

# use_roxygen_md() sets set up roxygen to use markdown syntax (bold, italics, lists, and links)
usethis::use_roxygen_md()

# install.packages() installs the styler package if you haven't already.
install.packages("styler")

# styler's style_pkg() applies styling to your entire package.
styler::style_pkg()

```

